---
title: "Lab 11: structure prediction with AlphaFold2 (structural bioinformatics pt2.)"
author: "Torrey Rhyne (A14397504)"
format: pdf
---

## The EBI AlphaFold database

Here we post process and inspect our modeling results from AlphaFold2.

My results from AF live in the folder/directory ``

```{r}
results_dir <- "HIVprdimer_23119/"
pdb_files <- list.files(results_dir, pattern = ".pdb", full.names = T)
```

We first ned to align and superpose these PDB models and we can use the `pdbaln()` function for this:
```{r}
library(bio3d)
pdbs <- pdbaln(pdb_files, fit = T, exefile = "msa")
```

## The RMSD matrix

A common measure of structural dis-similarity is called RMSD (root mean square distance):
```{r}
rd <- rmsd(pdbs, fit=T)
rd
```

Visualize:
```{r}
library(pheatmap)

rownames(rd) <- paste0("m",1:5)
colnames(rd) <- paste0("m",1:5)
pheatmap(rd)
```
Models 1/2 and 4/5 are similar.

Let's view in Mol*. Here we want the fitted coordinates.
```{r}
xyz <- pdbfit(pdbs, outpath = "fitted")
```

A full atom based fitting that our superposition did not work very well because we have multiple chains that are in different conformations.

I want to focus the superposition on the most invariant part (rigid "core").
```{r}
core <- core.find(pdbs)
core.inds <- core
```

```{r}
xyz <- pdbfit(pdbs, inds = core.inds, outpath = "core_fitted")
```

To evaluate how good multi-chain or multi-domain models are we need to looks at the PAE scores (predicted aligned error).

These are output as JSON format files. Let's find all their file names:
```{r}
pae_files <- list.files(results_dir, pattern = "0.json", full.names = T)
```

```{r}
library(jsonlite)

pae1 <- read_json(pae_files[1], simplifyVector = T)
pae5 <- read_json(pae_files[5], simplifyVector = T)

```

```{r}
attributes(pae1)
pae1$max_pae
pae5$max_pae
```

```{r}
plot.dmat(pae5$pae,
          xlab = "Residue No.",
          ylab = "Residue No.",
          zlim = c(0, 30))

plot.dmat(pae1$pae,
          xlab = "Residue No.",
          ylab = "Residue No.",
          zlim = c(0, 30))
```

## Main points
1. We can run AlphaFold on google compute (don't have to install ourselves). 
2. We can read these results into R and process to help us make sense of these models and their PAE and pLDDT scores.