---
title: "Class 19 Mini Project"
author: "Torrey Rhyne (A14397504)"
format: pdf
---

## Investigating pertussis cases by year

The United States Centers for Disease Control and Prevention (CDC) has been compiling reported pertussis case numbers since 1922 in their National Notifiable Diseases Surveillance System (NNDSS). 

We can view this data on the CDC website here: https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html

Setup:
```{r}
# install.packages("datapasta")
library(datapasta)
library(ggplot2)
library(jsonlite)
library(lubridate)
library(dplyr)
```

Q1. With the help of the **R “addin” package datapasta** assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

Paste CDC data:
```{r}
cdc <- data.frame(
                            year = c(1922L,1923L,1924L,1925L,1926L,1927L,
                                     1928L,1929L,1930L,1931L,1932L,1933L,1934L,
                                     1935L,1936L,1937L,1938L,1939L,1940L,
                                     1941L,1942L,1943L,1944L,1945L,1946L,
                                     1947L,1948L,1949L,1950L,1951L,1952L,
                                     1953L,1954L,1955L,1956L,1957L,1958L,
                                     1959L,1960L,1961L,1962L,1963L,1964L,
                                     1965L,1966L,1967L,1968L,1969L,1970L,
                                     1971L,1972L,1973L,1974L,1975L,1976L,1977L,
                                     1978L,1979L,1980L,1981L,1982L,1983L,
                                     1984L,1985L,1986L,1987L,1988L,1989L,
                                     1990L,1991L,1992L,1993L,1994L,1995L,
                                     1996L,1997L,1998L,1999L,2000L,2001L,
                                     2002L,2003L,2004L,2005L,2006L,2007L,
                                     2008L,2009L,2010L,2011L,2012L,2013L,
                                     2014L,2015L,2016L,2017L,2018L,2019L,2020L,
                                     2021L),
                            cases = c(107473,164191,165418,152003,202210,181411,
                                     161799,197371,166914,172559,215343,
                                     179135,265269,180518,147237,214652,
                                     227319,103188,183866,222202,191383,191890,
                                     109873,133792,109860,156517,74715,
                                     69479,120718,68687,45030,37129,60886,
                                     62786,31732,28295,32148,40005,14809,
                                     11468,17749,17135,13005,6799,7717,9718,
                                     4810,3285,4249,3036,3287,1759,2402,
                                     1738,1010,2177,2063,1623,1730,1248,
                                     1895,2463,2276,3589,4195,2823,3450,4157,
                                     4570,2719,4083,6586,4617,5137,7796,
                                     6564,7405,7298,7867,7580,9771,11647,
                                     25827,25616,15632,10454,13278,16858,
                                     27550,18719,48277,28639,32971,20762,
                                     17972,18975,15609,18617,6124,2116)
  )
```

Plot:
```{r}
p1 <- ggplot(cdc) +
  aes(x = year, y = cases) +
  geom_point() +
  geom_line() +
  labs(x = "Year", y = "Number of Cases", title = "CDC Pertussis Case Numbers Since 1922")
p1
```

## A tale of two vaccines (wP & aP)

Let’s return to our CDC data plot and examine what happened after the switch to the acellular pertussis (aP) vaccination program.

Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?
Also add the 2019 COVID pandemic.

```{r}
p2 <- p1 + 
  geom_vline(xintercept = 1946, color = "red") +
  geom_vline(xintercept = 1996, color = "blue") +
  geom_vline(xintercept = 2019, color = "darkgreen")
p2
```

Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

The pertussis field has several hypotheses for the resurgence of pertussis including: 
1) more sensitive PCR-based testing
2) vaccination hesitancy 
3) bacterial evolution (escape from vaccine immunity)
4) waning of immunity in adolescents originally primed as infants with the newer aP vaccine as compared to the older wP vaccine

## Exploring CMI-PB data

The CMI-PB project collects and makes available data on the immune response to Pertussis booster vaccination. 

Why is this vaccine-preventable disease on the upswing? To answer this question we need to investigate the mechanisms underlying waning protection against pertussis. This requires evaluation of pertussis-specific immune responses over time in wP and aP vaccinated individuals.

The new and ongoing CMI-PB project aims to provide the scientific community with this very information. In particular, CMI-PB tracks and makes freely available long-term humoral and cellular immune response data for a large number of individuals who received either DTwP or DTaP combination vaccines in infancy followed by Tdap booster vaccinations. This includes complete API access to longitudinal RNA-Seq, AB Titer, Olink, and live cell assay results directly from their website: https://www.cmi-pb.org/

**The CMI-PB API returns JSON data** (more robust than downloading a bunch of files)

Access the data via the API.
We will use the **jsonlite** package to access the data using the `read_json()` function.

```{r}
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE)
head(subject, 3)
```

Q4. How many aP and wP infancy vaccinated subjects are in the dataset?
```{r}
table(subject$infancy_vac)
```

Q5. How many Male and Female subjects/patients are in the dataset?
```{r}
table(subject$biological_sex)
```

Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?
```{r}
df <- data.frame(subject$race, subject$biological_sex)
table(df)
```

Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}

subject$age <- today() - ymd(subject$year_of_birth)

# I
subject_wP <- subject %>% filter(infancy_vac == "wP")
subject_wP$age <- today() - ymd(subject_wP$year_of_birth)
mean(subject_wP$age)/365
print("years, not days")

# II 
subject_aP <- subject %>% filter(infancy_vac == "aP")
subject_aP$age <- today() - ymd(subject_aP$year_of_birth)
mean(subject_aP$age)/365
print("years, not days")

# III
tt <- t.test(time_length(subject_wP$age, "years"),
              time_length(subject_aP$age, "years"))
tt$p.value
```

Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

Read the specimen and ab_titer tables into R and store the data as specimen and titer named data frames.

```{r}
# Complete the API URLs...
specimen <- read_json("https://www.cmi-pb.org/api//specimen", simplifyVector = TRUE) 
titer <- read_json("https://www.cmi-pb.org/api//plasma_ab_titer", simplifyVector = TRUE) 

head(specimen)
head(titer)
```

Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
meta <- inner_join(specimen, subject)
dim(meta)
head(meta)
```

Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta)
dim(abdata)

head(abdata)
```

Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

Number of rows (entries) has gone down.

## Examine IgG Ab titer levels

Now using our joined/merged/linked abdata dataset filter() for IgG isotype.

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
   xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

FIM2/3 vs. Fim2/3?

```{r}
oops <- abdata %>% filter(antigen=="Fim2/3")
table(oops$dataset)
```

Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?
FIM2/3, FHA, PT - These are pertussis proteins.

We can attempt to examine differences between wP and aP here by setting color and/or facet values of the plot to include infancy_vac status (see below). However these plots tend to be rather busy and thus hard to interpret easily.

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  xlim(0,75) +
  theme_bw()
```

Another version of this plot adding infancy_vac to the faceting:
```{r}
igg %>% filter(visit != 8) %>%
ggplot() +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  xlim(0,75) +
  facet_wrap(vars(infancy_vac, visit), nrow=2)
```

Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
PT <- igg %>% filter(antigen == "PT")

ggplot(PT) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE)

FIM <- igg %>% filter(antigen == "FIM2/3")

ggplot(FIM) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE)
```

Lets finish this section by looking at the 2021 dataset IgG PT antigen levels time-course:

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

Q18. Does this trend look similar for the 2020 dataset?

```{r}
abdata.20 <- abdata %>% filter(dataset == "2020_dataset")

abdata.20 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2020 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

