---
title: "class 8: Halloween project"
author: Torrey Rhyne (A14397504)
format: html
---

Import data:
```{r}
candy_file <- "candy-data.csv"
candy = read.csv(candy_file, row.names=1)
head(candy)


```

Q1. How many different candy types are in this dataset?
```{r}
nrow(candy)
```


Q2. How many fruity candy types are in the dataset?
```{r}
sum(candy$fruity)
```

Q3. What is your favorite candy in the dataset and what is it’s winpercent value?
```{r}
candy["Twix",]$winpercent
```

Q4. What is the winpercent value for “Kit Kat”?
```{r}
candy["Kit Kat",]$winpercent
```

Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?
```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

```{r}
# install.packages("skimr")
library("skimr")
skim(candy)
```

Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?
Yes, winpercent is on a different scale.

Q7. What do you think a zero and one represent for the candy$chocolate column?
1 = it is a chocolatey candy (TRUE), 0 = it's not a chocolatey candy (FALSE).

Q8. Plot a histogram of winpercent values
```{r}
library(ggplot2)
ggplot(candy, aes(x=winpercent)) + geom_histogram()
```


Q9. Is the distribution of winpercent values symmetrical? 
No.

Q10. Is the center of the distribution above or below 50%?
Below.

Q11. On average is chocolate candy higher or lower ranked than fruit candy?
```{r}
mean(candy$winpercent[as.logical(candy$nougat)])
mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruity)])

```

Chocolate candy is ranked higher.

Q12. Is this difference statistically significant?
```{r}
choc <- candy$winpercent[as.logical(candy$chocolate)]
fruit <- candy$winpercent[as.logical(candy$fruity)]
t.test(choc,fruit)
```

Yes, the difference is ranked higher.

Q13. What are the five least liked candy types in this set?
```{r}
head(candy[order(candy$winpercent), ], n =5)
```

Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, Jawbusters

Q14. What are the top 5 all time favorite candy types out of this set?
```{r}
head(candy[order(candy$winpercent, decreasing = T), ], n =5)
```

Reese's PB cups, Reese's miniatures, Twix, Kit Kat, Snickers.

Q15. Make a first barplot of candy ranking based on winpercent values.
```{r}
ggplot(candy) + aes(x=winpercent, y=rownames(candy)) + geom_col()
```

Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent? 
Also add useful color.
```{r}
colors = rep("black", nrow(candy))
colors[as.logical(candy$chocolate)] = "chocolate"
colors[as.logical(candy$bar)] = "brown"
colors[as.logical(candy$fruity)] = "pink"

ggplot(candy) + aes(winpercent, reorder(rownames(candy), winpercent)) + geom_col(fill= colors)

```

Now, for the first time, using this plot we can answer questions like:
Q17. What is the worst ranked chocolate candy?
Boston Baked Beans
Q18. What is the best ranked fruity candy?
Starburst

```{r}
# install.packages("ggrepel")
library(ggrepel)
```

Plot price vs. winpercent:
```{r}
ggplot(candy) + aes(winpercent, pricepercent, label = rownames(candy)) + geom_point(col=colors) 
  # + geom_text_repel(col=colors, size=3.3, max.overlaps = 5)
```

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```


Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck? 
Tootsie Roll Midgies

Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?
Nik L Nip (least popular), Nestle Smarties, Mr. Good Bar, Hershey's, Milk Chocolate.

Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().
```{r}
# base plot
ggplot(candy, aes(pricepercent, reorder(rownames(candy), winpercent))) + geom_col()
# make a lollipop chart
ggplot(candy, aes(pricepercent, reorder(rownames(candy), pricepercent))) + geom_segment(aes(yend=reorder(rownames(candy), pricepercent), xend = 0), col = "gray40") + geom_point()
```

```{r}
# install.packages("corrplot)
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
chocolate and fruity

Q23. Similarly, what two variables are most positively correlated?
```{r}
cor(candy$chocolate, candy$winpercent)
cor(candy$chocolate, candy$bar)
```

winpercent and chocolate

PCA:
```{r}
pca <- prcomp(candy, scale = T)
summary(pca)
```

Plot:
```{r}
plot(pca$x[,1:2], col=colors, pch=16)
```

Combine data and plot:
```{r}
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
  aes(x=PC1, y=PC2, size=winpercent/100, text = rownames(my_data), label = rownames(my_data)) + geom_point(col=colors)
p
```

```{r}
p + # + geom_text_repel(size=3.3, col=colors, max.overlaps = 7)
 theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
# install.packages("plotly")
library(plotly)
ggplotly(p)
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? 
Pluribus and fruity. Yes, this makes sense. Most of the fruity candies are on the right side (positive PC1 value) of the previous plot. Fruity candies usually come in a bag or box of multiple (unlike chocolate bars).
