---
title: "Class 9: Unsupervised Learning Mini-Project"
author: Torrey Rhyne (A14397504)
format: pdf
---

## Exploratory data analysis

Download and import data:
```{r}
# save your input data file into your project directory
fna.data <- "WisconsinCancer.csv"
# input the data and store as wisc.df
wisc.df <- read.csv(fna.data, row.names=1)

# view
head(wisc.df)

# remove the first column
wisc.data <- wisc.df[,-1]

# view 
head(wisc.data)

# Create diagnosis vector for later 
diagnosis <- wisc.df[,1]
diagnosis <- factor(diagnosis)
```

Familiarize yourself with the data:

Q1. How many observations are in this dataset?
```{r}
nrow(wisc.data)
```

Q2. How many of the observations have a malignant diagnosis?
```{r}
table(diagnosis)
```

Q3. How many variables/features in the data are suffixed with _mean?
```{r}
grep("_mean",colnames(wisc.data))
```

## Principal Component Analysis

Does the data need to be scaled?
```{r}
# check column means and standard deviations
colMeans(wisc.data)
# apply
apply(wisc.data,2,sd)
```

PCA
```{r}
# Perform PCA on wisc.data 
wisc.pr <- prcomp(wisc.data, scale = T)
# Look at summary of results
summary(wisc.pr)
```

Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)? 44.27%
Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data? 3 (cumulative proportion of PC3 = 72.64%)
```{r}
v <- summary(wisc.pr)
pcvar <- v$importance[3,]
which(pcvar >= 0.7)[1]
```

Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data? 7 (cumulative proportion of PC7 = 91.01%)

Interpreting PCA results

```{r}
biplot(wisc.pr)
```

Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?
This plot is very difficult to interpret. All the points / labels overlap.

Make it look nicer:
```{r}
# Scatter plot observations by components 1 and 2
plot(wisc.pr$x[, 1:2], col = diagnosis,
     xlab = "PC1", ylab = "PC2")

# Scatter plot observations by components 1 and 2
plot(wisc.pr$x[, c(1,3)], col = diagnosis, 
     xlab = "PC1", ylab = "PC2")

# note: had to save diagnosis vector as factor
```

Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?
The PC1 vs. PC2 and PC1 vs. PC3 plots do a nice job at separating the malignant / benign diagnoses. But since PC2 captures more of the variance than PC3, the first plot is better.

Use ggplot to make a better figure:
```{r}
# Create a data.frame for ggplot
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis

# Load the ggplot2 package
library(ggplot2)

# Make a scatter plot colored by diagnosis
ggplot(df) + 
  aes(PC1, PC2, col=diagnosis) + 
  geom_point()
```

Variance explained.

```{r}
# calculate variance
pr.var <- wisc.pr$sdev^2
head(pr.var)

# Variance explained by each principal component: pve
pve <- pr.var/ sum(pr.var)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")

# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )

## ggplot based graph
#install.packages("factoextra")
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

Communicating PCA results.

Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC.
```{r}
wisc.pr$rotation["concave.points_mean",1]
```

The feature "concave points" negatively contributes to PC1 (for our plots with PC1 on the x-axis, this pushes points to the left = more likely to be malignant).

## Hierarchial clustering

Go back to wisc.data and scale:
```{r}
data.scaled <- scale(wisc.data)
```

Calculate (Euclidean) distance between points:
```{r}
data.dist <- dist(data.scaled)
```

hclust:
```{r}
wisc.hclust <- hclust(data.dist, method= "complete")
```

Interpreting the results.

Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?
```{r}
plot(wisc.hclust)
# cut the tree
abline(h=19, col="red", lty=2)
```

Selecting number of clusters.

Cut tree to get 4 clusters:
```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, h = 19)
```

Compare cluster membership to diagnosis:
```{r}
table(wisc.hclust.clusters, diagnosis)
```

We should probably use a different number of clusters so that the patients are slightly more evenly distributed (cluster 4 only has 2 patients).

Using different methods.

Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.
```{r}
# complete
wisc.hclust <- hclust(data.dist, method= "complete")
plot(wisc.hclust)
# single
wisc.hclust <- hclust(data.dist, method= "single")
plot(wisc.hclust)
# average
wisc.hclust <- hclust(data.dist, method= "average")
plot(wisc.hclust)
# ward.D2
wisc.hclust <- hclust(data.dist, method= "ward.D2")
plot(wisc.hclust)
```

I think the plot for "ward.D2" looks the best. Let's try cutting the tree to into 2 clusters and see how well separated the malignant / benign diagnoses are.
```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, h = 55)
table(wisc.hclust.clusters, diagnosis)
```
This is pretty good!

## Combining Methods

Clustering on PCA results. Let’s see if PCA improves or degrades the performance of hierarchical clustering.

```{r}
# hierarchial clustering with ward.D2 method, cut in 2 clusters
wisc.pr.hclust <- hclust(data.dist, method= "ward.D2")
grps <- cutree(wisc.pr.hclust, k=2)

# examine
table(grps)
table(grps, diagnosis)

# plot
plot(wisc.pr$x[,1:2], col=grps)
plot(wisc.pr$x[,1:2], col=diagnosis)
# make colors match up
g <- as.factor(grps)
levels(g)
g <- relevel(g,2)
levels(g)
# Plot using our re-ordered factor 
plot(wisc.pr$x[,1:2], col=g)
```

Visualize in 3D:
```{r}
# install.packages("rgl")
library(rgl)
plot3d(wisc.pr$x[,1:3], xlab="PC 1", ylab="PC 2", zlab="PC 3", cex=1.5, size=1, type="s", col=grps)
# rglwidget(width = 400, height = 400)
```

Cluster the PCA data (distance along the first 7 PCs)
```{r}
# calculate distance
dist <- dist(wisc.pr$x[,1:7])
# hclust
wisc.pr.hclust <- hclust(dist, method="ward.D2")
# cut into 2 clusters
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
# examine results
table(wisc.pr.hclust.clusters)
```

Q13. How well does the newly created model with four clusters separate out the two diagnoses?
```{r}
# table
table(wisc.pr.hclust.clusters, diagnosis)
# what proportion were correctly classified?
(329+188)/length(diagnosis)
```

91% correct is pretty good! (but context matters, this is a cancer diagnosis)

Q14. How well do the hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? 
```{r}
# hclust with method = "complete" and 4 clusters
wisc.hclust <- hclust(data.dist, method= "complete")
wisc.hclust.clusters <- cutree(wisc.hclust, h = 19)
# table
table(wisc.hclust.clusters, diagnosis)
# what proportion were correctly classified?
(343+165)/length(diagnosis)

# hclust with method = "ward.D2" and 2 clusters
wisc.hclust <- hclust(data.dist, method= "ward.D2")
wisc.hclust.clusters <- cutree(wisc.hclust, k=2)
# table
table(wisc.hclust.clusters, diagnosis)
# what proportion were correctly classified?
(337+164)/length(diagnosis)
```

The original clustering is still very good! 

Q15. OPTIONAL: Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?
I think clustering PCA data is the best by a small margin.

## Prediction

Use of PCA model from before to predict new data
```{r}
# data
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc

# plot
plot(wisc.pr$x[,1:2], col=g)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

Q16. Which of these new patients should we prioritize for follow up based on your results?
We should follow up with patient 2 (more likely to be malignant)

```{r}
sessionInfo()
```

